Basics of cubic spline models

### 2021-11-27 text/21/cubic-splines.Rmd
Added open-access-deal.md

@@

This is the matrix of independent variables for a single cubic polynomial.

```{r}
kable(xm[1:21, 1:4])
```

@@

The trick to fitting four separate cubic polynomials is to "restart" the cubic terms after x=5, x=10, and x=15.

```{r}
kable(xm[1:20, ])
```

@@

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Continuous but not smooth cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Continuous and smooth (1st derivative) cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Continous and smooth (1st and 2nd derivatives) cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
kable(cubic_spline[1:21, ])
```

@@

```{r}
linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
kable(linear_spline[1:21, ])
```

@@

You can probably see the linearity below 5 and above 15.

>>>>

This is what the various pieces look like numerically ...

```{r}
kable(xm[1:21, 1:16])
```

... and graphically.

@@

@@

The trick to fitting four separate cubic polynomials is to "restart" the intercept, linear, quadratic, and cubic terms after x=5, x=10, and x=15, as shown above. This leads to a model with 16 degrees of freedom. This is way too many degrees of freedom for only 20 data points, but it helps anchor a series of more reasonable models.

@@

```{r}
lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Continuous but not smooth cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Continuous and smooth (1st derivative) cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Continous and smooth (1st and 2nd derivatives) cubic splines") %>%
    display_dots(simulated_example)
```

@@

```{r}
cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
```

@@

```{r}
linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
```

@@

You can probably see the linearity at the extremes of the data.
